\begin{center}
    {\large \textbf{ACKNOWLEDGEMENTS}}
\end{center}

I would like to express my deepest gratitude to my advisor, mentor, and committee chair, Dr. P. Richard Hahn, whose guidance, support, and expertise have been instrumental throughout my research journey. His depth of knowledge in Bayesian statistics and tree-based models has shaped the foundation of this work. As both a mentor and a friend, his example is one I aspire to emulate in my own career.

I am sincerely thankful to my committee members, Dr. Shuang Zhou and Dr. Shiwei Lan, for their flexibility, insightful feedback, and thoughtful contributions, all of which have enriched the depth and rigor of this thesis. In particular, I would like to thank Dr. Zhou for her inspiring teaching and intellectual curiosity, which played a pivotal role in both my undergraduate and master's success. 

Special thanks to the School of Mathematical and Statistical Sciences at Arizona State University for fostering an environment of academic excellence and growth. I am especially grateful for the sense of community and support that the School provides—an environment that has shaped not only my academic path but also my personal development.

I am deeply grateful to my father for instilling in me a lifelong intellectual curiosity and passion for learning, and for being the earliest source of encouragement in pursuing mathematics and statistics. My family unwavering support, especially during periods of serious illness has been foundational to every step of my academic journey. I am especially indebted to my mother for her steadfast love, consistency, and strength during those most difficult times. Her support made my success possible. 

Lastly, I acknowledge the researchers whose foundational work on low birth weight determinants and Bayesian nonparametric methods laid the groundwork for this study. This research stands on their shoulders.
\clearpage